---
title: "Working Class Representative Performance"
subtitle: "Which factors positively contribute to votes in their favor"
format: html
editor: visual
execute:
  echo: true
---

# **Introduction and Data**

The data set came from a journal published at Political Research Quarterly (2023). The authors were Eric Hansen, who is the assistant professor of political science at Loyola University Chicago and Sarah A. Treul, an Associate Professor of political science at UNC.

The authors collected, recorded, and analyzed novel data describing the occupational background of all candidates who competed in the U.S. House primaries between 2008 and 2016.

The dataset was used to give us an idea of how well candidates perform in U.S. House primary elections. In total, there are 7869 observations and 24 variables. Each observation represents a candidate who competed in the U.S. Hour primaries between 2008 and 2016. Each row of observation included a variety of factors related to election results, such as vote share, fundraising, incumbency status, education level, ideology, number of opponents, and political affiliation. Not all variables are useful in terms of predicting the candidate's performance as some of them are only numerical descriptions of vote share.

This research question is important for the following reasons. Given that the working class is less represented among high-level elected government officials, we would like to encourage research dedicated to this issue in order to identify factors that contribute to this phenomenon. Research in this area has focused on the barriers hindering working-class representatives. As a result, we do not know much about what factors allow working-class elected officials to overcome these barriers to election and thus, what factors influence their positive performance. We have decided to focus our research on working-class officials who have been elected to office as a case study. Results from our research would provide valuable insight to future working-class candidates. Ultimately, we hope this project can provide insight to future working class candidates providing more guidance to their campaign strategies and fundraising efforts. As a result, we hope this could contribute to more equality and equity in the government.

# **Literature**

**Primary Barriers to Working Class Representation**

Treul, S. A., & Hansen, E. R. (2023). Primary Barriers to Working Class Representation. Political Research Quarterly, 0(0). https://doi.org/10.1177/10659129231154914

This research article is written by Sarah Treul who co-published the dataset Primary Barriers to Working Class Representation which we hope to utilize in our research. Her research seeks to answer how working class candidates fare in primary elections given that working class candidates are a minority on the ballot compared to white-collar candidates. The research article analyzed occupational backgrounds of primary election candidates for the U.S. House of Representatives from 2008 to 2016 with election results. Treul et al. found that from 2008 to 2016, working class candidates are 31% less likely to win elections and tend to score 24% less votes than white collar candidates. In their analysis, Treul et al. suggests that lower success in primary elections for the working class may be due lower funding and negative bias among party leaders towards working class candidates. 

This article uses the Primary Barriers to Working Class Representation dataset to analyze election results based on whether the candidates were working class and then analyzes factors that affected why working class candidates did not win elections. Our research would seek to understand how working class candidates overcame barriers to winning primary elections by analyzing factors that were outliers among candidates that did win the election. 

# **Methodology**

***Variables Selection:***

1.  Dependent variable: **candpct**, which represents the candidate's vote share in the primary election.

2.  Independent variables: We select the variables that potentially measure factors related to the candidate's vote share. These selected 14 independent variables include:

-   state_abbrev (the state the candidate is from)

-   candnumber (count of opponents in the candidate's primary election)

-   cand_party (the party of the candidate)

-   quality_cand (0=candidate is nonquality nonincumbent, 1=candidate is quality nonincumbent, 2=candidate is incumbent)

-   qualnumber (count of quality opponents)

-   openseat (whether the incumbent is running for reelection)

-   partyDonors1 (count of donations candidate received from donors to their party's national campaign funds)

-   college (percentage of the congressional district's residents who hold four-year college degrees)

-   dist_income (median income of the congressional district)

-   total_primary (total dollars contributed toward a candidate's campaign funds and outside spending, unit: millions of dollars)

-   third (whether candidate runs as a third party or independent)

-   gop (whether the candidate is Republican)

-   ideo_extreme (absolute value of Bonica's CFscore measure of candidate ideology)

-   former_worker (whether the candidate used to be a worker but is no longer now)

***Analysis Procedure:***

1.  First, we filter the dataset and keep only candidates with working-class background (worker == 1) and have won the election (winwin == 1).

2.  Second, we use the forward and backward model selection method to compare and choose the variables that best predict the dependent variable. Specifically, for the forward selection, we first try to fit each variable in a linear regression model with a single independent variable and find out which fits the best (the one with the lowest AIC). Then, we try to add each variable into the existing best model, and find out the best. Theoretically, we repeat this process by trying to add more variables and compare the AIC until we find out the best model. But since the number of variables is too large, we use the stepAIC method in R to perform the forward model selection. To see which model selection methods work better and generate more useful results, we also use the stepAIC method to perform the backward model selection. Based on the results, we select the model with the lowest AIC and conclude the most prominent factors related to the vote share as well as the coefficients of the factors.

3.  Then, using the factors selected, we also run an interactive multiple linear regression model to see which of the additive and interactive model works better. Based on the results, we select the model with the lower AIC.

4.  Lastly, to get make the results more straightforward and readable, we visualize the results by \[Note: we are confused about this part and we left a comment at the end of the Results section. We might need some help with this part, thanks!\]

# **Results**

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
library(tidymodels)
library(MASS)
library(ggplot2)

```

```{r}
#| label: load-data-2
candidate <- read.table("data/Treul_Hansen_Replication_Data.tab", 
                        sep = "\t", header = TRUE)
write.csv(candidate, file = "data/candidate.csv", col.names = TRUE)

glimpse(candidate)
```

```{r}
#| label: filter-candidate 

candidate_won <- candidate |>
  filter(winwin == 1, worker == 1) #337 observations

candidate_won_new <- candidate_won[,-c(2, 4, 5, 6, 9, 10, 22, 23)]
  
candidate_final <- candidate_won_new |>
  drop_na()
```

```{r}
#| label: count-observations 

candidate_final |>
  nrow() 
```



```{r}
#| label: backward-selection 


lm1 <- lm(candpct ~ state_abbrev + candnumber + cand_party + quality_cand
          + qualnumber + openseat + PartyDonors1 + college + dist_income
          + total_primary + third + gop + former_worker, 
          data = candidate_final)
stepAIC(lm1,  direction = "backward") 

```

```{r}
#| label: forward-selection 


lm2 <- lm(candpct ~ state_abbrev, data = candidate_final)

stepAIC(lm2,  scope = ~ state_abbrev + candnumber + cand_party + quality_cand
          + qualnumber + openseat + PartyDonors1 + college + dist_income
          + total_primary + third + gop + former_worker,
        direction = "forward")

```

**selected model**

lm(formula = candpct \~ candnumber + quality_cand + qualnumber + openseat + dist_income + total_primary, data = candidate_final)

This is the model selected through backward selection. It has the lowest AIC and fewer variables than the best model generated through forward selection.

we also picked the top three variables: candnumber, dist_income, total_primary

```{r}
#| label: interaction 

lmInt <- lm(formula = candpct ~ candnumber * quality_cand * qualnumber * openseat * dist_income * total_primary, data = candidate_final)


linear_reg()|>
  set_engine("lm") |>
  fit(candpct ~ candnumber * quality_cand * qualnumber * openseat * 
        dist_income * total_primary, data = candidate_final) |>
  glance() |>
  pull(AIC)

```

conclusion: the interactive multiple linear regression model gave us a worse AIC, -447, which was higher than 1236.62. Therefore, we will keep the original addictive model.

```{r}
#| label: dataset-4-visualization 

candidate_final <- candidate_final |>
  mutate(candnumber2 = if_else(candnumber < 4, "small pool of candidates", 
                               "large pool of candidates")) |>
  mutate(primary2 = if_else(total_primary == 0.00000, "NA", "Other"))

```

```{r}
#| label: summary-stas-for-variables-of-interest 

  
```

```{r}
#| label: visualization 

candidate_final |>
  ggplot(aes(x = dist_income, y = candpct, color = total_primary)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "orange", 
                       breaks = 
                         c(0, 0.005, 0.1, 0.5, 1.0, 1.5)) +
  geom_point(data = subset(candidate_final, primary2 == "NA"), size = 0.4, 
             color = "grey") +
  geom_smooth(method = lm, se = FALSE, color = "black") +
  facet_wrap(~ candnumber2) +
  theme_classic() +
  labs(x = "median income of the congressional district (100k)", 
       y = "vote share in the primary election (%)", 
       color = "campaign funds\n& outside spending\n(million$)", 
       title = "Primary predictors to working-class candiates vote share",
       subtitle = "among candidates that won the primary elections", 
       caption = "**points marked grey are those with 0$
       campaign funds & outside spending as indicated in the dataset"
       ) 





```

